\mbox{Verifying Imperative Higher-Order Programs}

### 2009-10-09 papers/esop2010/main.tex
go

The first obstacle has a well known solution using a type and effect
system~\cite{talpin94taed} with effect variables: effect polymorphism does
exactly what we want here. Using effect polymorphism, we can say that {\tt f}
(the argument of {\tt iter}) has type $α->^e\unitml$, whera $α$ is the type of
the array elements and where $e$ is an {\em effect variable}, which amounts to
say that {\tt f} has a {\em certain} effect $e$. The complete type of {\tt
iter} is thus:
\begin{displaymath}
  ∀α∀e. (α ->^e \unitml) ->^\emptyset α~\mathtt{array} ->^e \unitml
\end{displaymath}
The total effect of {\tt iter} is then $e$, meaning {\tt iter}
modifies whatever {\tt f} modifies.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Two-state
postconditions are a standard way to get rid of most auxiliary variables.
However, the pre- and postcondition of {\tt f} may not refer to arbitrary
parts of the state, but only the part {\tt f} reads or modifies.  With these
conventions, one can now write a generic implementation and specification for
the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
of our prototype \who. There are a few syntactic differences with \ocaml, in
particular for arrays: the type of arrays is {\tt 'a farray ref}, the length
of the array {\tt a} is {\tt length !a} in the program and {\tt length !!a} in
the logic, and similarily access to the {\tt i}th element is {\tt get i !a} in
the program and {\tt get i !!a} in the logic.

@@

The specification of {\tt iter} is correct, but it will be difficult to prove
in general, because it is not clear wether {\tt f} modifies the array or not,
or more precisely, if the effect variable {\tt e} does contain {\tt a} or not.
This is an instance of the {\em aliasing problem} (see, for
example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
is not well-suited for programs with aliasing. Without special techniques (for
example {\em separation logic}), a large number of equalities and disequalities
are needed to reason about potentially aliased references.

@@

The store types $\efft{ε}$ describe a portion of the store and are basically
mappings from region variables to values. The domain of the mapping is the
effect $ε$, where effect variables stand for unknown parts of the domain. To
dea with objects of these types, we introduce three symbols
\begin{eqnarray*}
  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
\end{eqnarray*}
The first one denotes a combining operation for stores; the resulting store
contains all the mappings of the second store, as well as all the mappings of
the first one, as long as they are not present in the second one. The second
operations is indexed by an effect and it restricts a store to a certain
effect. Finally, the double bang {\tt !!} is the logic equivalent to the
lookup operator {\tt !}. It is pure and it is always with respect to a certain
state.

@@

The specification of {\tt memo} is very simple. The only precondition is that
the table should only contain bindings $y |-> {\mathtt f}~y$, which is
expressend by the {\tt stores} predicate. The postcondition states that the
result is equal to {\tt f0~x} and that the table still verifies the {\tt
stores} predicate.

>>>>

The first obstacle has a well known solution using a type and effect
system~\cite{talpin94taed} with effect variables: effect polymorphism does
exactly what we want here. Using effect polymorphism, we can say that {\tt f}
(the argument of {\tt iter}) has type $α->^e\unitml$, where $α$ is the type of
the array elements and where $e$ is an {\em effect variable}, which amounts to
say that {\tt f} has a {\em certain} effect $e$. The complete type of {\tt
iter} is thus:
\begin{displaymath}
  ∀α∀e. (α ->^e \unitml) ->^\emptyset α~\mathtt{array} ->^e \unitml
\end{displaymath}
The total effect of {\tt iter} is then $e$, meaning {\tt iter}
modifies whatever {\tt f} modifies.

@@

The main idea of this paper is to combine the effect-based Hoare logic of
\cite{Filliatre00a} with the more powerful effect system of
\cite{talpin94taed} and the more powerful annotations of
\cite{regis-gianas-pottier-08} to obtain a modular system for proving
higher-order programs with side effects. As in~\cite{regis-gianas-pottier-08},
we use operators \preml\ and \postml\ to access the pre- and postcondition of
a function. However, in our system, a precondition takes as an argument not
only the function argument, but also the initial state, before the execution
of the function. The postcondition takes the argument and return value, but
also the states before {\em and} after the function call. Two-state
postconditions are a standard way to get rid of most auxiliary variables.
However, the pre- and postcondition of {\tt f} may not refer to arbitrary
parts of the state, but only the part {\tt f} reads or modifies.  With these
conventions, one can now write a generic implementation and specification for
the {\tt iter} function. It is given~Fig.~\ref{fig:iter}, in the input syntax
of our prototype \who. There are a few syntactic differences with \ocaml, in
particular for arrays: the type of arrays is {\tt 'a farray ref}, the length
of the array {\tt a} is {\tt length !a} in the program and {\tt length !!a} in
the logic, and similarly access to the {\tt i}th element is {\tt get i !a} in
the program and {\tt get i !!a} in the logic.

@@

The specification of {\tt iter} is correct, but it will be difficult to prove
in general, because it is not clear whether {\tt f} modifies the array or not,
or more precisely, if the effect variable {\tt e} does contain {\tt a} or not.
This is an instance of the {\em aliasing problem} (see, for
example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
is not well-suited for programs with aliasing. Without special techniques (for
example {\em separation logic}), a large number of equalities and disequalities
are needed to reason about potentially aliased references.

@@

The store types $\efft{ε}$ describe a portion of the store and are basically
mappings from region variables to values. The domain of the mapping is the
effect $ε$, where effect variables stand for unknown parts of the domain. To
deal with objects of these types, we introduce three symbols
\begin{eqnarray*}
  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
\end{eqnarray*}
The first one denotes a combining operation for stores; the resulting store
contains all the mappings of the second store, as well as all the mappings of
the first one, as long as they are not present in the second one. The second
operations is indexed by an effect and it restricts a store to a certain
effect. Finally, the double bang {\tt !!} is the logic equivalent to the
lookup operator {\tt !}. It is pure and it is always with respect to a certain
state.

@@

The specification of {\tt memo} is very simple. The only precondition is that
the table should only contain bindings $y |-> {\mathtt f}~y$, which is
expressed by the {\tt stores} predicate. The postcondition states that the
result is equal to {\tt f0~x} and that the table still verifies the {\tt
stores} predicate.
